---
title: "Your Paper Title Here"
author: "Arnav, Maxx, Mohit"
date: Date
output: pdf_document
urlcolor: blue
header-includes:
- \usepackage{amsmath,amsfonts,amssymb}
- \usepackage{setspace} \doublespacing
fontsize: 11pt
---

```{r setup, include=FALSE}
#Use this code chunk to include libraries, and set global options.
```

# Abstract
Write your abstract here.

# Keywords
*Keyword 1*, *Keyword 2*, *Keyword 3*, *Keyword 4*

\newpage

# Introduction / Background
Provide an introduction with background information.

```{r label="important_R_code", eval=TRUE, echo=FALSE}
# You can reference your code in the appendix (sample here).
```

# Statistical Methods

## Model
Describe the statistical model used.

$T$ is the random variable that defines the number of people who had the vaccine out of the people who tested positive of a total $n = 170$ cases. An 
\begin{align*}
    T &\sim Binom(n = 170, \pi) \\
    \pi &= P(\text{Vaccinated} | \text{Positive Covid Test}) \\
    \pi &= \frac{n_1 \pi_v}{n_1 \pi_v + n_2 \pi_p}, \text{Where $n_1 \approx n_2$, the randomization is $1:1$} \\
    \pi_v &= P( \text{Testing positive } | \text{Having vaccine }) \\
    \pi_p &= P( \text{Testing positive } | \text{Having placebo }) \\
    \pi &\frac{\pi_v}{\pi_v + pi_p} \\
    \psi &= 1 - \frac{\pi_v}{\pi_p} \\
    \psi &= \frac{1 - 2\pi}{1 - \pi} \\
    \pi &= \frac{1 - \psi}{2 - \psi} \\
    \pi_v &= \frac{8}{17411} \\
    \pi_p &= \frac{162}{17511} \\
    \psi &\approx 0.9503 \\
    \pi &\approx 0.0473 \\
    T &\sim Binom(n=170, \pi = 0.0473)
\end{align*}


We observed a $t$ of $8$, and the parameter of interest is $\psi = \frac{1-2\pi}{1-\pi}$

We are interested in the interval estimate of $\psi$, our $H_0: \psi = 0.3$ and our $H_1: \psi \neq 0.3$

We are calculating two different estimators for $\psi$, using maximum likelihood as well as bayes.


## Likelihood Inference
Detail the likelihood approach.

\begin{align*}
    T &\sim Binom(170, \pi_0) \leftrightarrow T_1, T_2, \dots, T_{170} \sim Bernouli(\pi_0) \\
    \hat\pi_0^{MLE} &= \frac{T}{n} \approx 0.0473 \\
    \hat\psi_0^{MLE} &= \frac{1 - 2\pi_0}{1-\pi_0} = \frac{n-2T}{n-T} = \frac{170 - 2(8)}{170-8} \approx 0.9506
\end{align*}

Standard errors and significance tests are not invariant to transformations, which means the likelihood function will be written in terms of $\psi$


\begin{align*}
    g(\psi) &= \frac{1-\psi}{2-\psi} \\
    L^*(\psi) &= L(g(\psi)) = (^n_t) (\frac{1-\psi}{2-\psi})^t (1 - \frac{1-\psi}{2-\psi})^{n-t} \\
    \ell^*(\psi) &= log(L^*(\psi)) = log(^n_t) + t \cdot log(\frac{1-\psi}{2-\psi}) + (n-t)log(1 - \frac{1-\psi}{2-\psi}) \\
    \frac{d}{d\psi}\ell^*(\psi) &= \frac{n-t}{2-\psi} - \frac{t}{(1-\psi)(2-\psi)} \\
    0 &= \frac{n-t}{2-\psi} - \frac{t}{(1-\psi)(2-\psi)}, \text{For MLE condition} \\
    \hat\psi_0^{MLE} &\approx 0.9506 \\
    &\text{Second derivative test} \\
    \frac{d^2}{d\psi^2} \ell^* (\psi) &= \frac{n-t}{(2-\psi)^2} + \frac{t(2\psi - 3)}{((1-\psi)(2-\psi))^2} \\
    \frac{d^2}{d\psi^2} \ell^* (0.9506) &\approx -3123.83 \\
    &\text{Thus function concave downwards, meaning global maximum} \\
    \hat\psi_0^{MLE} &\approx Norm(\psi_0, \sqrt{\frac{1}{n I(\psi_0)}}) \\
    I(\psi_0) &= E [ \frac{-d^2}{d\psi_0^2} log(f_{\psi_0}(T))] \\
    f_{\psi_0}(T) &= (^n_t) \pi^t (1-\pi)^{n-t}
\end{align*}

## Bayesian Inference
Detail the Bayesian approach.

# Results
Present your findings. 

# Discussion / Conclusion
Discuss / conclude here.

# Bibliography

Brown, B. (2024). *Lecture Title*. Lecture slides, Course Name, University Name.

Doe, J. (2020). Title of the Paper. *Journal Name*, 12(3), 45-67.

Last, F., & Last, F. (2025). *Book Title*. Publisher.

Smith, A., & Johnson, C. (2023). *Title of the Online Article*. Retrieved from https://www.example.com.

# Appendix

## Code
```{r ref.label = "important_R_code", eval=FALSE}
```

## Proofs
If applicable, include detailed mathematical derivations or additional theoretical explanations.
