---
title: "Your Paper Title Here"
author: "Arnav, Maxx, Mohit"
date: Date
output: pdf_document
urlcolor: blue
header-includes:
- \usepackage{amsmath,amsfonts,amssymb}
- \usepackage{setspace} \doublespacing
fontsize: 11pt
---

```{r setup, include=FALSE}
#Use this code chunk to include libraries, and set global options.
```

# Abstract
Write your abstract here.

# Keywords
*Keyword 1*, *Keyword 2*, *Keyword 3*, *Keyword 4*

\newpage

# Introduction / Background
Provide an introduction with background information.

```{r label="important_R_code", eval=TRUE, echo=FALSE}
# You can reference your code in the appendix (sample here).
```

# Statistical Methods

## Model
Describe the statistical model used.

$T$ is the random variable that defines the number of people who had the vaccine out of the people who tested positive of a total $n = 170$ cases. An 
\begin{align*}
    T &\sim Binom(n = 170, \pi) \\
    \pi &= P(\text{Vaccinated} | \text{Positive Covid Test}) \\
    \pi &= \frac{n_1 \pi_v}{n_1 \pi_v + n_2 \pi_p}, \text{Where $n_1 \approx n_2$, the randomization is $1:1$} \\
    \pi_v &= P( \text{Testing positive } | \text{Having vaccine }) \\
    \pi_p &= P( \text{Testing positive } | \text{Having placebo }) \\
    \pi &\frac{\pi_v}{\pi_v + pi_p} \\
    \psi &= 1 - \frac{\pi_v}{\pi_p} \\
    \psi &= \frac{1 - 2\pi}{1 - \pi} \\
    \pi &= \frac{1 - \psi}{2 - \psi} \\
    \pi_v &= \frac{8}{17411} \\
    \pi_p &= \frac{162}{17511} \\
    \psi &\approx 0.9503 \\
    \pi &\approx 0.0473 \\
    T &\sim Binom(n=170, \pi = 0.0473)
\end{align*}


We observed a $t$ of $8$, and the parameter of interest is $\psi = \frac{1-2\pi}{1-\pi}$

We are interested in the interval estimate of $\psi$, our $H_0: \psi = 0.3$ and our $H_1: \psi \neq 0.3$

We are calculating two different estimators for $\psi$, using maximum likelihood as well as bayes.


## Likelihood Inference
Detail the likelihood approach.

\begin{align*}
    T &\sim Binom(170, \pi_0) \leftrightarrow T_1, T_2, \dots, T_{170} \sim Bernouli(\pi_0) \\
    \hat\pi_0^{MLE} &= \frac{T}{n} \approx 0.0473 \\
    \hat\psi_0^{MLE} &= \frac{1 - 2\pi_0}{1-\pi_0} = \frac{n-2T}{n-T} = \frac{170 - 2(8)}{170-8} \approx 0.9506
\end{align*}


Standard errors and significance tests are not invariant to transformations, which means the likelihood function will be written in terms of $\psi$

\begin{align*}
    g(\psi) &= \frac{1-\psi}{2-\psi} \\
    L^*(\psi) &= L(g(\psi)) = (^n_t) (\frac{1-\psi}{2-\psi})^t (1 - \frac{1-\psi}{2-\psi})^{n-t} \\
    \ell^*(\psi) &= log(L^*(\psi)) = log(^n_t) + t \cdot log(\frac{1-\psi}{2-\psi}) + (n-t)log(1 - \frac{1-\psi}{2-\psi}) \\
    \frac{d}{d\psi}\ell^*(\psi) &= \frac{n-t}{2-\psi} - \frac{t}{(1-\psi)(2-\psi)} \\
    0 &= \frac{n-t}{2-\psi} - \frac{t}{(1-\psi)(2-\psi)}, \text{For MLE condition} \\
    \hat\psi_0^{MLE} &\approx 0.9506 \\
    &\text{Second derivative test} \\
    \frac{d^2}{d\psi^2} \ell^* (\psi) &= \frac{n-t}{(2-\psi)^2} + \frac{t(2\psi - 3)}{((1-\psi)(2-\psi))^2} \\
    \frac{d^2}{d\psi^2} \ell^* (0.9506) &\approx -3123.83 \\
    &\text{Thus function concave downwards, meaning global maximum} \\
    \hat\psi_0^{MLE} &\approx Norm(\psi_0, \sqrt{\frac{1}{n I(\psi_0)}}) \\
    I(\psi_0) &= E [ \frac{-d^2}{d\psi_0^2} log(f_{\psi_0}(T))] \\
    f_{\psi_0}(T) &= (^n_t) \pi^t (1-\pi)^{n-t} \\
    f_{\psi_0}(T) &= (^n_t) (\frac{1-\psi}{2-\psi})^t (1-\frac{1-\psi}{2-\psi})^{n-t} = L^*(\psi) \\
    log(f_{\psi_0}(T)) &= \ell^*(\psi) \\
    I(\psi_0) &= E[-(\frac{n-t}{(2-\psi)^2} + \frac{t(2\psi - 3)}{((1-\psi)(2-\psi))^2})] \\
    I(\psi_0 = 0.9506) &= 3123.8335, \text{plugging in our known values}\\
    \sqrt{\frac{1}{nI(\psi_0)}} &= 0.00137 \\
    CI_{\alpha=0.05}(\hat\psi_0^{MLE}) &= 0.9506 \pm 1.96 \cdot 0.00137 \\
    CI_{\alpha=0.05}(\hat\psi_0^{MLE}) &= [0.9479, 0.9533]
\end{align*}

Using likelihood inference, we are 95 percent confident the true value of $\psi_0$ is in the interval: $[0.9479, 0.9533]$


### Taylor Plot

```{r, warning=FALSE}
library(fastR2)
loglikelihood <- function(psi){
  n <- 170
  t <- 8
  return(log(choose(n, t)) + t * log((1 - psi) / (2 - psi)) + (n - t) * log(1 - (1 - psi) / (2 - psi)))
}

ml.psi <- maxLik2(loglik = loglikelihood, start=0.3)
plot(ml.psi) + labs(title = "Taylor Approximation Plot")
```


### Bootstrap Confidence Interval

```{r}
set.seed(123)

n <- 170
psi_mle <- 0.9506
pi_hat <- (1 - psi_mle) / (2 - psi_mle)

n_bootstrap <- 10000

T_bootstrap <- rbinom(n_bootstrap, size = n, prob = pi_hat)

psi_bootstrap <- (n - 2 * T_bootstrap ) / (n - T_bootstrap)

ci_boot <- quantile(psi_bootstrap, probs = c(0.025, 0.975))

print(ci_boot)
```

\begin{align*}
    H_0&: \psi_0 = 0.3 \\
    H_1&: \psi_0 \neq 0.3 \\
    w &= 2log(\Lambda) \\
    \Lambda &= \frac{L(\hat\psi_0^{MLE})}{L(\psi_{null})} \\
    w &= 121.72
\end{align*}

```{r}
p_value <- pchisq(121.72, df=1, lower.tail=FALSE)
```

Our p-value is: `r p_value`, which is significantly lower than our significance value of $\alpha = 0.05$, so as a result we reject the null as there is convincing evidence that the true value for $\psi \neq 0.3$, which means that the efficacy of the vaccine is different than the FDA requirement of 30 percent

```{r}
set.seed(414)
psinull <- 0.3
B <- 10000
n <- 15

sim_func <- function(i){
  xstar <- rweibull( n = 15, shape = 2, scale = 1/sqrt(psinull * pi))
  
  psihatstar <- n / (pi * sum(xstar^2))
  wstar <- 2 * (loglikelihood(psihatstar) - loglikelihood(psinull))
  return(wstar)
}

null_sim <- lapply(1:B, sim_func)

Wstar <- c(unlist(null_sim))
emp_value <- sum(Wstar >= 121.72) / B
print(emp_value)
```

## Bayesian Inference
Detail the Bayesian approach.

```{r}
set.seed(123)

samples <- 10000
a <- 8 + 43.03
b <- 170 - 8 + 43.03

library(tidyverse)

ggplot() +
  geom_function(fun = dbeta, 
                mapping = aes(color = 'Prior'),
                args = list(shape1 = 43.03, shape = 43.03),
                xlim = c(0, 1))+
  geom_function(fun = dbeta,
                mapping = aes(color = 'Posterior'),
                args = list(shape1 = a, shape2 = b),
                xlim = c(0, 1)) +
  scale_color_manual(name = 'dist', values = c('blue', 'red'))

pi_median <- qbeta(0.5, a, b)
print(pi_median)
pi_ci <- qbeta(p = c(0.025, 0.975), a, b)

psi_median <- (1 - 2 * pi_median) / (1 - pi_median)

library(HDInterval)
hd_pi_ci <- hdi(qbeta, credMass=0.95, shape1 = a, shape2 = b)
pi_lower <- unname(hd_pi_ci[1])
pi_upper <- unname(hd_pi_ci[2])
psi_upper <- (1 - 2 * pi_lower) / (1 - pi_lower)
psi_lower <- (1 - 2 * pi_upper) / (1 - pi_upper)
psi_lower
psi_upper
```

\begin{align*}
    P(\frac{1-\psi}{2-\psi} < 0.1985|t ) &= 0.5 \\
    P(1-\psi < 0.1985(2-\psi)|t) &= 0.5\\
    P(\psi > 0.7523 | t) &= 0.5
\end{align*}

Therefore the posterior median of $\psi$ is 0.7523

```{r}
psi_median_p_value <- 2 * pbeta(0.7523, a, b, lower.tail = FALSE) # for symmetrical two sided test
```

Our p-value for $\psi$ is less than 0.05 from our bayesian inference, therefore we have convincing evidence that the true value for the efficacy of the vaccine is different than the 30 percent FDA requirement. Our 95% confidence interval for $\psi$ does not contain 0.3 and only has values larger, which provides evidence that this vaccine exceeds the FDA efficacy requirement


# Results
Present your findings. 

# Discussion / Conclusion
Discuss / conclude here.

# Bibliography

Brown, B. (2024). *Lecture Title*. Lecture slides, Course Name, University Name.

Doe, J. (2020). Title of the Paper. *Journal Name*, 12(3), 45-67.

Last, F., & Last, F. (2025). *Book Title*. Publisher.

Smith, A., & Johnson, C. (2023). *Title of the Online Article*. Retrieved from https://www.example.com.

# Appendix

## Code
```{r ref.label = "important_R_code", eval=FALSE}
```

## Proofs
If applicable, include detailed mathematical derivations or additional theoretical explanations.
